Formal Modelling of a Microcontroller Instruction Set in B

### 2009-06-27 trunk/papers/SBMF2009/source/paper.tex
Mais algumas retificações dos revisores realizadas.

\documentclass[a4paper]{llncs}

@@

@@

The B method~\cite{Abrial} supports the construction of safety systems
model by verification of proofs that guarantees its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined until the implementation model. Development
environments based on the B method also include source code generators
for several programming languages, but the result of this translation
cannot be compared by formal means. \cite{Dantas_SBMF08} presented
recently an approach to extend the scope of the B method up to the
assembly level language language. One key component of this approach
is to build, within the framework of the B method, formal modelos of
the instruction set of such assembly languages.

@@

Other possible uses of a formal model of a micro-controller
instruction set include documentation, the construction of simulators,
and be possibly the starting point of a verification effort for the
actual implementation of a Z80 design. Moreover the model of the
instruction set could be instrumented with non-functional aspects,
such as the number of cycles it takes to execute an instruction, to
prove lower and upper bounds on the execution time of a routine. The
goal of this project, though, is to provide a basis for the generation
of software artifacts at the assembly level that are amenable to 
refinement verification within the B method.

@@

The B method for software development~\cite{Abrial} is based on the B
Abstract Machine Notation (AMN) and the use of formally proved
refinements up to a specification sufficiently concrete that
programming code can be automatically generated from it. Its
mathematical basis consists of first order logic, integer arithmetic
and set theory, and its corresponding constructs are very similar to
those of the Z notation

A B specification is structured in modules. A module defines a set of
valid states, including a set of initial states, and operations that
may provoke a transition between states. The design process starts
with a module with a so-called functional model of the system under
development. In this initial modelling stage, the B method requires
that the user proves that, in a machine, all the initial states are
valid, and that operations do not define transitions from valid states
to invalid states.

@@

We have been developed a reusable set of basic definitions to model
hardware concepts and data types concepts. These definitions are
grouped into two separate development projects and are available as
libraries. A third project is devoted to the higher-level aspects of
the Z80 model. Thus, the workspace is composed of: a hardware library,
a types library and a project for the specific platform, in this case
the Z80. The corresponding dependency diagram is depicted in
Figure~\ref{fig:hardware-definition-graph}; information specific to
each project is presented in the following.

\begin{figure}[h] \centering
\includegraphics[width=.92\textwidth]{diagramaEstrutural_vertical.jpg}
 \caption{Dependency diagram of the Z80 model.}
\label{fig:hardware-definition-graph}
\end{figure}

\subsection{Bit representation and manipulation}

@@

The $\mathit{BYTE\_INDEX}$ is the domain of the functions modelling
bytes. It starts at 1 to obey a definition of sequences from B
method. However, it is common in hardware architectures to start
indexing from zero. The definition $\mathit{PHYS\_BYTE\_INDEX}$ is
used to provide functionalities obeying this convention. The
$\mathit{BYTE}$ type is a specialized type from
$\mathit{BIT\_VECTOR}$, but it has a size limit. Other specific
definitions are provided to facilitate further modelling: the type
$\mathit{BV16}$ is created for bit vector of length 16 in a similar
way.

@@

The \textit{Z80} is a CISC microcontroller developed by
\textit{Zilog}~\cite{Z80_manual}. It supports 158 different
instructions. These instructions are classified into these categories:
load and exchange; block transfer and search; arithmetic and logical;
rotate and shift; bit manipulation (set, reset, test); jump, call and
return; input/output; and basic cpu control.

@@

\subsection{Modelling registers, input and output ports and instructions}

The Z80 has different types of registers and instructions. The CPU
contains general-purpose registers ($\mathit{id\_reg\_8}$), a stack
pointer ($\mathit{sp}$), program counter ($\mathit{pc}$), two index
registers ($\mathit{ix}$ and $\mathit{iy}$), an interrupt register
($\mathit{i\_}$), a refresh register ($\mathit{r\_}$), two bits
($\mathit{iff1}$, $\mathit{iff2}$) used to control the interruptions,
a pair of bits to define the interruption mode ($\mathit{im}$) and the
input and output ports ($\mathit{i\_o\_ports}$). Below, part of the
corresponding definitions are replicated from the \textit{INVARIANT}:

\begin{sloppypar}
\bf INVARIANT
\hspace*{0.10in}\it rgs8  $\in$  \it id\_reg\_8  $\fun$  \it BYTE  $\land$ pc  $\in$  \it INSTRUCTION  $\land$

@@

The microcontroller model can specify several security properties. For example, the last operation could have a
restriction to write only in a defined region of memory.

@@

However, few proof commands\footnote{The proof commands are steps that
  direct the prover to find the proof, and cannot introduce false
  hypothesis.} need to be used to prove most proof obligations. As
there are many similar assembly instructions, some human-directed
proofs, when replayed, could discharge other proof obligations. A good
example is a set of 17 proof commands that quickly aided the
verification of 99\% (2295) of proofs WD. Finally, all of the 2926
proof obligations were proved using the tool support of the
development environment.

@@

There are in the literature of computer science some
approaches~\cite{BHDL_2003,GEMPLUS_99} to model hardware and the
virtual machines using the B method. Then, the B method has been used
successfully to model the operational semantic. However the cost of
modelling was still expensive and this paper quoted some techniques to
lower the cost of modelling. 
% Although, the modelling of the Z80 do not
%have modelled some small aspects.

@@

In general, the researchers employing the B method have focused on
more abstract level of description of software.  Considering low-level
aspect, there has been previous work on modelling the Java Virtual
Machine\cite{GEMPLUS_99}.

@@

This work has shown an approach to the formal modelling of the
instruction set of microcontrollers using the B method. During the
construction of this model, several ambiguities and errors were
encountered in the during the official reference~\cite{Z80_manual}. As
the B notation has a syntax that is not too distant from that of
imperative programming languages, such model could be used to improve
the documentation used by assembler programmers. Besides, the formal
notation used is analyzed by software that guarantees the
correctness of typing, the well-definedness of expressions, in addition
to safety properties of the micro-controller state.

@@

\bibitem {BHDL_2003}    
Aljer, P. Devienne, S. Tison and J-L. Boulanger andG. Mariano. Bhdl: Circuit Design in B. A. In
ACSD, Third International Conference on Application of Concurrency to System Design, pages 241-242, 2003.

@@

\bibitem {Dantas_SBMF08}
Dantas, B; D\'{e}harbe, D; Galv\~{a}o, S; Et al. Applying the B Method to Take on the Grand Challenge of
Verified Compilation. In: SBMF, Savaldor, 2008. SBC.

@@

\bibitem{Z80_manual}
Zilog. Z80 Family CPU User Manual. www.zilog.com/docs/z80/um0080.pdf

>>>>

\documentclass{llncs}

@@

\hyphenation{se-pa-ra-te}

@@

The B method~\cite{Abrial} supports the construction of safety systems
model by verification of proofs that guarantees its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined until the implementation model. Development
environments based on the B method also include source code generators
for programming languages, but the result of this translation
cannot be compared by formal means. The paper \cite{Dantas_SBMF08} presented
recently an approach to extend the scope of the B method up to the
assembly level language. One key component of this approach
is to build, within the framework of the B method, formal models of
the instruction set of such assembly languages.

@@

Other possible uses of a formal model of a microcontroller
instruction set include documentation, the construction of simulators,
and be possibly the starting point of a verification effort for the
actual implementation of a Z80 design. Moreover the model of the
instruction set could be instrumented with non-functional aspects,
such as the number of cycles it takes to execute an instruction, to
prove lower and upper bounds on the execution time of a routine. The
goal of this project, though, is to provide a basis for the generation
of software artifacts at the assembly level that are amenable to 
refinement verification within the B method.

@@

The B method for software development~\cite{Abrial} is based on the B Abstract
Machine Notation (AMN) and the use of formally proved refinements up to a
specification sufficiently concrete that programming code can be automatically
generated from it. Its mathematical basis consists of first order logic, integer
arithmetic and set theory, and its corresponding constructs are similar to those
of the Z notation.

A B specification is structured in modules. A module defines a set of
valid states, including a set of initial states, and operations that
may provoke a transition between states. The design process starts
with a module with a so-called functional model of the system under
development. In this initial modelling stage, the B method requires
that the user proves that, in a machine, all the its initial states are
valid, and that operations do not define transitions from valid states
to invalid states.

@@

We have been developed a reusable set of basic definitions to model
hardware concepts and data types concepts. These definitions are
grouped into two separate development projects and are available as
libraries. A third project is devoted to the higher-level aspects of
the platform. Thus, the workspace is composed of: a hardware library,
a types library and a project for the specific platform, in this case
the Z80. The corresponding dependency diagram is depicted in
Figure~\ref{fig:hardware-definition-graph}; information specific to
each project is presented in the following.

\begin{figure}[h] \centering
\includegraphics[width=.92\textwidth]{diagramaEstrutural_vertical2.png}
 \caption{Dependency diagram of the Z80 model.}
\label{fig:hardware-definition-graph}
\end{figure}

\subsection{Bit Representation and Manipulation}

@@

The $\mathit{BYTE\_INDEX}$ is the domain of the functions modelling bytes. It
starts at 1 to obey a definition of sequences from B. However, it is common in
hardware architectures to start indexing from zero. The definition
$\mathit{PHYS\_BYTE\_INDEX}$ is used to provide functionalities obeying this
convention. The $\mathit{BYTE}$ type is a specialized type from
$\mathit{BIT\_VECTOR}$, but it has a size limit. Other specific definitions are
provided to facilitate further modelling: the type $\mathit{BV16}$ is created for
bit vector of length 16 in a similar way.

@@

The \textit{Z80} is a CISC microcontroller developed by
\textit{Zilog}~\cite{Z80_manual}. It supports 158 different instructions and all
of them were specified. These instructions are classified into these categories:
load and exchange; block transfer and search; arithmetic and logical; rotate and
shift; bit manipulation; jump, call and return; input/output; and basic cpu
control.

@@

\subsection{Modelling Registers, Input and Output Ports and Instructions}

The Z80 has different types of registers and instructions. The CPU
contains general-purpose registers ($\mathit{id\_reg\_8}$), a stack
pointer ($\mathit{sp}$), program counter ($\mathit{pc}$), two index
registers ($\mathit{ix}$ and $\mathit{iy}$), an interrupt register
($\mathit{i\_}$), a refresh register ($\mathit{r\_}$), two bits
($\mathit{iff1}$, $\mathit{iff2}$) used to control the interruptions,
a pair of bits to define the interruption mode ($\mathit{im}$) and the
input and output ports ($\mathit{i\_o\_ports}$). Below, part of the
corresponding definitions are replicated from the $\INVARIANT$:

\begin{sloppypar}
%\bf INVARIANT
\hspace*{0.10in}\it rgs8  $\in$  \it id\_reg\_8  $\fun$  \it BYTE  $\land$ pc  $\in$  \it INSTRUCTION  $\land$

@@

The microcontroller model can specify security properties. For example, the
last operation could have a restriction to write only in a defined region of memory.

@@

However, few proof commands\footnote{The proof commands are steps that direct the
prover to find the proof, and cannot introduce false hypothesis.} need to be used
to prove most proof obligations. As there are many similar assembly instructions,
some human-directed proofs, when replayed, could discharge other proof
obligations. A good example is a set of 17 proof commands that quickly aided the
verification of 99\% (2295) of proofs WD. A new feature of B environment also
was used, the parallelization. This used four computers each one with two cores and
it increased many times the capacity of data processing. Finally, all of the
2926 proof obligations were proved using the tool support of the development
environment.

@@

There are in the literature of computer science some
approaches~\cite{BHDL_2003,GEMPLUS_99} to model hardware and the virtual machines
using the B method. Then, the B method has been used successfully to model the
operational semantic. However the cost of modelling was still expensive and this
paper quoted some techniques to lower the cost of modelling.
% Although, the modelling of the Z80 do not have modelled some small aspects.

@@

In general, the researchers employing the B method have focused on
more abstract level of description of software.  Considering low-level
aspect, there has been previous work on modelling the Java Virtual
Machine~\cite{GEMPLUS_99}.

@@

This work has shown an approach to the formal modelling of the
instruction set of microcontrollers using the B method. During the
construction of this model, some ambiguities and errors were
encountered in the official reference~\cite{Z80_manual}. As
the B notation has a syntax that is not too distant from that of
imperative programming languages, such model could be used to improve
the documentation used by assembler programmers. Besides, the formal
notation used is analyzed by software that guarantees the
correctness of typing, the well-definedness of expressions, in addition
to safety properties of the microcontroller state.

@@

\bibitem {BHDL_2003}    
Aljer, P. Devienne; S. Tison  J-L. Boulanger and G. Mariano. Bhdl: Circuit Design in B. A. In
ACSD, Third International Conference on Application of Concurrency to System Design, pages 241-242, 2003.

@@

\bibitem {Dantas_SBMF08}
Dantas, B; D\'{e}harbe, D.; Galv\~{a}o, S. L.; Moreira, A. M. and
Medeiros Jr, V. G.. Applying the B Method to Take on the Grand Challenge of
Verified Compilation. In: SBMF, Savaldor, 2008. SBC.

@@

\bibitem{Z80_manual}
Zilog. Z80 Family CPU User Manual. http://www.zilog.com/docs/z80/um0080.pdf
